\bf Estimating the number of clusters using Cross Validation

### 2015-12-05 New simulation/technometrics.tex
Add new directory contains new simulation

@@

Although the condition given by Proposition $2$ is succinct in such that it's straight forward to see how
the structure of covariance matrix $\Sigma_{XY}$ affects the performance of Gabriel CV method, it's not simple and easy to interpret such condition as the one in Proposition $1$. In specific, each block of matrix $\Sigma_{XY}$ influences the condition while the combined effect
determines whether Gabriel CV method chooses $k=1$ or $k=2$.

In practice, check such condition is tricky. One need to know the covariance matrix of each cluster to check if the condition holds. However, in order to estimate the covariance matrix we need to know the cluster center of each cluster, which in turn need to know the number of clusters in the data. A possible solution is following
\begin{enumerate}
	\item For parameter $k$ starts at $k=1$ do the following
	\begin{enumerate}[label=(\roman*)]
		\item Estimate $k$ cluster centers by applying $k$-means with parameter $k$
		\item Estimate the covariance matrix $\hat{\Sigma}$ for each cluster 
		\item Check the condition for each $\hat{\Sigma}$
		\item If the the condition doesn't hold, rotate that cluster through 
		rotation matrix $R^T$, where $\hat{\Sigma} = RLR^T$ 
		\item Apply previously proposed Gabriel CV method on the rotated data, get estimated number $k^*$
	\end{enumerate}
	\item If $k^* = k$, stop the algorithm and return $k$; otherwise repeat step $1$ for parameter $k+1$
\end{enumerate}
from which we can see the condition checking step can not be separated from the proposed Gabriel CV method.
Another empirically observed solution is to leave out most columns for clustering, i.e. increase the 
dimension of $\Sigma_{YY}$. Such method is observed reducing the impact of correlation on the performance 
of the proposed method. Although the theory for such solution is available for special cases,
its general version hasn't been developed.

@@

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected. The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.  }
\label{fig:overlap-color_plot}
\end{figure}
%----------------------------------------------------------------
\section{Simulation}
In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. We select $p=q$ and $5-$fold cross-validation in row
($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
column.

\begin{enumerate}
  \item Single cluster in $10$ dimensions --- $200$ observations, each
    observation uniformly distributed over $[0,1]$ in each dimension. 
  \item Two clusters in $4$ dimensions --- $50$ i.i.d
    observations are generated from both multivariate normal
    $\mathcal{N}\left(\mu_1,0.5\Sigma\right)$ and multivariate normal
    $\mathcal{N}\left(\mu_2,1.5\Sigma\right)$, where 
    cluster center $\mu_1 = (1,0,0,1)$ and $\mu_2 =(1,3.5,3.5,1)$
    respectively. $\Sigma$ has $AR(1)$ structure with $\rho=-0.2$ and $\sigma=1$.
  \item Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$
    i.i.d standard normal observations, with cluster centers randomly
    generated from multivariate normal distribution
    $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$
  \item Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$
    i.i.d standard normal observations, with cluster centers randomly
    generated from multivariate normal distribution
    $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$
  \item Four log-normal clusters in $16$ dimensions --- For each cluster, $30$
    or $60$ i.i.d centered log-normal observations are generated from
    $ln\mathcal{N}\left(0,0.5^2 \right)$. The cluster centers are randomly
    generated from multivariate normal distribution
    $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$
  \item[6.] Three exponential clusters in $20$ dimensions with different
    variance --- $40$ observations in each cluster are generated from centered
    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
    with $\boldsymbol\mu$ randomly generated from
    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
    1/5$ respectively.
\end{enumerate}
Note that in setting $2$--$6$, all clusters are well-separated, i.e. no
overlapping. In fact, any simulated clusters with minimum distance less than
1 unit was discarded, so there is clear definition of true number of
clusters. The parameters in setting $2$--$6$ are chosen such that about half of
the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}.

Table \ref{table1} shows the distribution of $k$ selected by each algorithm in 
each simulation setting. We run $k$ from $1$ to $15$, and the best $k$ within this range
is selected by the algorithms. NA in the table means algorithm did not converge or otherwise 
failed. 
\clearpage

\begin{center}
\footnotesize
\begin{longtable}{lcccccccccccccccc}
\captionsetup{justification=centering}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
  {\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\midrule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & NA \\
\midrule
\endhead
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{\footnotesize
  \textsuperscript{$\ast$} CH, Hartigan, and Stability are excluded from
  Setting~1 because they cannot select $k=1$.
}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
\end{center}

\noindent
All methods are used with their default parameter setting except for Gap
statistics. We selected the $k$  corresponds to the global maximum of gap
statistics, rather than the default method which selects the smallest $k$ such
that the gap statistics is not more than 1 standard error away from the first
local maximum. This is because among all the possible options, the global
maximum criteria gives the best result in our simulation settings.

For CH-index, Hartigen and Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
included for comparison since the true $k=1$. It also demonstrates the difficulty 
that null scenario poses for some existing algorithms. For those could select $k=1$,
most of them work well under such setting except for Jump method. Setting $2-4$ have Gaussian
distribution, where the Gaussian mixture model-based method suppose to
perform the best. However, it only performs reasonably well in setting $2$. In setting $3$ and $4$,
it has a very poor performance even if the underlying distribution is indeed
Gaussian mixture model. It highlights the difficulty that high dimension
data could cause. Other methods (CH-index, Hartigan, Jump
method, Prediction strength and Boostrap stability) also have problems in
these two high-dimension settings as we could not find any one with reasonable
performance in terms of finding the correct $k$ in both cases. The only exception is
Gap statistics, which works perfectly in both settings. Our proposed Gabriel method (also Wold
CV method) clearly stands out from the basket of existing methods under the 
high-dimension setting $3$ and $4$, as we can see they both perform perfectly under
such settings.

Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
instead of normal distributed, coupled with unequal number of observations or
unequal variance for different clusters. From Table \ref{table1}, we can see none of 
the existing methods has good performance in both settings. In contrast,
our proposed Gabriel method is robust and has superior performance
compare with other methods. Because one can hardly tell the underlying
distribution of data in practice, the resilience to non-Gaussian data gives
the Gabriel method a clear edge. 
%----------------------------------------------------------------
\section{Real data application}
We also applied our proposed method to three real world data sets obtained
from the University of California Irvine machine learning repository. The
first and third data sets are selected because there are clear number of
clusters in those two data sets. The second data set is used as a benchmark
data set since it was widely used in literature.

>>>>

Among all the proposed methods, very few provide theoretical consistency result for choosing $k$. 
Assume data is a mixture of $G$ $p$-dimension clusters with equal prior, identically distributed with 
common covariance matrix and finite fourth moments in each dimension, \citet{sugar2003finding} shows
the Jump method will pick the correct $k$ under the conditions that cluster centers are sufficiently
separated and an appropriate transformation is used. \cite{tibshirani2005cluster} proved that the prediction 
strength method is consistent under the setting that the $k$ clusters are uniformly distributed on $k$ unit balls
in $p$-dimension, with pairwise distance of centers are at least four.

@@

Above equation gives the condition of when Gabriel CV method would correctly choose $k=1$ over $k=2$. Although the expression is succinct, it's not straight forward to see how the structure of covariance matrix $\Sigma_{XY} \in \mathbb{R}^{(p+q) \times (p+q)}$ affects the performance of Gabriel CV method. Here, assuming covariance matrix has compound symmetric structure where only the matrix dimension $p+q$ and $\rho$ are variables, i.e. $$\Sigma_{XY} = \begin{pmatrix}
1 & \rho & \cdots & \rho \\
\rho & 1 & \cdots & \rho \\
\vdots & \vdots & \cdots & \rho \\
\rho& \rho & \cdots & 1 \\
\end{pmatrix}
$$
we are able to feel what above equation implies for this specific case. Another reason to use the compound symmetric structure covariance matrix is that it's invariant under the permutation of each column vector in the matrix. Given that Gabriel CV method randomly choose $p$ columns as $X$ ($q$ columns as $Y$), compound symmetric structure insures that the covariance matrix $\Sigma_{XY}$ always look the same no matter which $p$ columns selected as $X$.

If we do $2$ fold cross-validation in the column, i.e. $p=q$, then the result in Proposition $2$ implies that the boundary value of $\rho$ is 
\[
	\rho^* = \frac{1}{p+1}
\]
under the compound symmetric structure given above, where $\Sigma_{XY}$ has $p+q=2p$ dimension. If $\rho < \rho^*$, then Gabriel CV prefers $CV(1)$ over $CV(2)$ and vice versa. It means the boundary value $\rho^*$ depends on the dimension of $\Sigma_{XY}$ linearly.

If dimension is $p+q = 2$ with $p=q$, then above boundary condition shows the boundary value $\rho^* = 0.5$, while for dimension $p+q = 100$, the  boundary value $\rho^* = \frac{1}{51}$. However, such value also depends on the ratio of $p$ over $q$. Also in $p+q = 100$ dimension, if we pick $p=2$ and $q=98$ (leave majority of the columns for clustering), the the boundary value $\rho^* \in (\frac{1}{6},\frac{1}{7})$.

@@

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected. The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.  }
\label{fig:overlap-color_plot}
\end{figure}

\subsection{Correlation adjustment for Gabriel method}
When the correlation between dimensions are high, the proposed Gabriel CV method tends to overestimate the number $k$. A simply remedy for the high correlation is available if we assume common covariance structure among the $k$ clusters. 
\begin{enumerate}
	\item Apply Gabriel CV method on the original data $\dataX$, get estimated number of cluster $\hat{k}$
	\item Estimate the pooled covariance matrix $\hat{\Sigma}$ from the $\hat{k}$ clusters.
	\item Let $\hat{\Sigma} = \Gamma\Lambda\Gamma'$ be the Eigendecomposition of $\hat{\Sigma}$, we rescale and rotate the original data $\dataX$ to get $ \widetilde{\dataX} = \dataX\Gamma\Lambda^{-1/2}Q$, where $Q$ is a random  orthonormal rotation matrix. 
	\item Apply Gabriel CV method again on the transformed data $ \widetilde{\dataX}$ to estimate $k$. 
\end{enumerate}

\section{Simulation}
In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. We select $p=q$ and $5-$fold cross-validation in row
($m=\frac{1}{4}n$) as default parameter setting for our proposed Gabriel
method. Note that set $p=q$ corresponding to $2-$fold cross-validation in
column.

\begin{enumerate}

  \item \textbf{Correlation between dimensions} -- Six clusters in $10$ dimensions. 
  	Each cluster has $100$ or $50$
    multivariate normal observations with common covariance matrix $\Sigma$ 
    which has compound symmetric structure with $1$ in diagonal and $\rho$ 
    off diagonal. For each $\rho$, $100$ data are simulated and the proportion each
    method successfully picks $k=6$ is reported. $\rho$ takes value in $\{0,0.1,...,0.9\}$.

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting1/Overlay.pdf}
	\label{fig:setting1}
	\end{figure}

  \item \textbf{Noise dimensions} --- Three clusters in $6$ dimensions. Each cluster has $100$ or $50$
    mean zero multivariate normal observations with identity covariance matrix. 
    We add $P$ dimensions of noise to the data, which is randomly generated from uniform$[0,1]$. The noise
    dimension $P$ takes values in $\{0,6,...,54\}$.

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting2/Overlay.pdf}
	\label{fig:setting2}
	\end{figure}

  \item \textbf{High dimension} --- Eight clusters in $p$ dimensions, $p$ takes values 
  	in $\{10,20,...,100\}$. Each cluster has $100$ or $50$ mean zero multivariate normal 
  	observations with identity covariance matrix.

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting3/Overlay.pdf}
	\label{fig:setting3}
	\end{figure}

     \item \textbf{Variance heterogeneity} --- Three clusters in $20$ dimensions, each has
     $60$ observations in it. Observations are generated from 
     $\mathcal{N}\left(\mathbf{0},\sigma_1^2\mathbf{I}\right)$,
      $\mathcal{N}\left(\mathbf{0},\sigma_2^2\mathbf{I}\right)$ and
       $\mathcal{N}\left(\mathbf{0},\sigma_3^2\mathbf{I}\right)$ where 
       $\sigma_1^2 : \sigma_2^2: \sigma_3^2 = 1:\frac{1+R}{2}:R$. The maximum ratio $R$
       takes values in $\{1,5,10,...,45\}$

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting4/Overlay.pdf}
	\label{fig:setting4}
	\end{figure}

	\item \textbf{Heavy tail data} --- Five clusters in $15$ dimensions, each has
     $80$ observations in it. Observations have independent $t$ distribution in each dimension  
     with degree of freedom $\nu$, which takes value in $\{2,3,...,11\}$

	\begin{figure}[H]
	\centering
	\includegraphics[width=5.5in, height=3.3in]{demo/bench/setting5/Overlay.pdf}
	\label{fig:setting4}
	\end{figure}
\end{enumerate}

Note that in all settings above, cluster centers are randomly generated from 
multivariate normal distribution $\mathcal{N}\left(\mathbf{0},\varsigma \mathbf{I}\right)$. 
All clusters are well-separated, i.e. no overlapping. In fact, any simulated 
clusters with minimum distance less than $1$ unit was discarded, so there is
clear definition of true number of clusters. The parameters $\varsigma$ is chosen 
such that about half of the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}.

\section{Real data application}
We also applied our proposed method to three real world data sets obtained
from the University of California Irvine machine learning repository. The
first and third data sets are selected because there are clear number of
clusters in those two data sets. The second data set is used as a benchmark
data set since it was widely used in literature.
